---
title: "Programming Assignment 3"
author: "Kirill Avilenko"
date: "10/1/2020"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Data exploration

### Outcome of care measures
The Outcome of Care Measures.csv table contains forty seven (47) fields. This table provides each hospital’s risk-adjusted 30-Day Death (mortality) and 30-Day Readmission category and rate.

Read data from given CSV file:
```{r}
outcome <- read.csv("outcome-of-care-measures.csv", colClasses = "character")
```



### Hospitals



## Part 1: Plot 30-Days mortality rates from heart attack

In order to plot 30-days mortality rates from heart attack we need to find and prepare data from "outcome-of-care-measures" table. The required info is given in column 11, but it is stored as a character.

```{r}
str(outcome[,11])
```

Convert data to numeric format and check data:
```{r}
outcome[, 11] <- as.numeric(outcome[, 11]);
str(outcome[,11])
```

Plot the histogram for 30-days mortality rate from heart attack:
```{r}
hist(outcome[, 11], 
     main='Hospital 30-Day Death (Mortality) Rates from Heart Attack', 
     xlab='Deaths', 
     col='red')
```

## Part 2: Best hospital in a state

In order to define best hospital in a state we need to design a function that takes two arguments:
* 2-character state code
* Outcome name - "heart attack", "heart failure" and "pneumonia"
and returns hospital name.

```{r}
best <- function(state, outcome) {
    ## Read outcome data
    outcomeTbl <- read.csv("outcome-of-care-measures.csv", colClasses = "character")

    ## Check that state is valid
    if (!state %in% unique(outcomeTbl[['State']])) {
        stop('invalid state') ## Throw an error with information message
    }
    ## Check if outcome is valid
    if (!outcome %in% c('heart attack', 'heart failure', 'pneumonia')) {
        stop('invalid outcome')
    }
    
    ## Return hospital name in that state with lowest 30-day death ## rate
}
```

## Ranking hospitals by outcome in a state

## Ranking hospitals in all states